Variations on a Bayesian Theme: Comparing Bayesian Models of Referential Reasoning

### 2014-03-14 02_RSA.tex
minor corrections
several footnotes for reviewers' comments

Referential games are confined interactive reasoning tasks used to
study pragmatic inference in a controlled environment \cite{Frank}. A
referential game consists of a set of objects called a
\textit{context}. A context contains different shapes of different
colors in various arrangements, such as shown in Figure~\ref{context}.
The speaker in the game is asked to refer to one of the objects (the
\emph{target}) by uttering an expression (typically a single word
denoting a feature, i.e., color or shape, of the target) to the
listener. The listener, who has does not know which object is the
target, needs to recover the target based on the speaker's choice of
expression.

@@

The speaker acts sub-optimally by choosing the utterance that soft-maximizes her expected utility, which is defined as the informativeness of the utterance subtracted by its cost:
\begin{equation} \label{Bayesian_sender_softmax}
\sigma(m \mid t) \propto \exp(\lambda_\mathrm{S}\cdot \mbox{U}(m,t))=\exp(\lambda_\mathrm{S} \cdot (\mbox{Info}(m,t)-\mbox{Cost}(m))),
\end{equation}
where $\lambda_\mathrm{S}$ is a parameter measuring the speaker's
degree of rationality, i.e., to what extent the speaker sticks to the
strict optimum. The cost term is used to encode preference in
different utterances, be it about the utterances' lengths or syntactic
categories.

From (\ref{Bayesian_literal})-(\ref{Bayesian_sender_softmax}) we obtain the speaker's production rule:

>>>>

Referential games are confined interactive reasoning tasks used to
study pragmatic inference in a controlled environment \cite{Frank}. A
referential game consists of a set of objects called a
\textit{context}. A context contains different shapes of different
colors in various arrangements, such as shown in Figure~\ref{context}.
The speaker in the game is asked to refer to one of the objects (the
\emph{target}) by uttering an expression (typically a single word
denoting a feature, i.e., color or shape, of the target) to the
listener. The listener, who does not know which object is the
target, needs to recover the target based on the speaker's choice of
expression.

@@

The speaker acts sub-optimally by choosing the utterance that soft-maximizes her expected utility, which is defined as the informativeness of the utterance subtracted by its cost:
\begin{equation} \label{Bayesian_sender_softmax}
\sigma(m \mid t) \propto \exp(\lambda_\mathrm{S}\cdot \mbox{U}(m,t))=\exp(\lambda_\mathrm{S} \cdot (\mbox{Info}(m,t)-\mbox{Cost}(m))),
\end{equation}
where $\lambda_\mathrm{S}$ is a parameter measuring the speaker's
degree of rationality, i.e., to what extent the speaker sticks to the
strict optimum\footnote{Mathematically, it can be shown that when $\lambda_\mathrm{S}=0$, $\sigma(m \mid t)$ is uniformly distributed over all messages, meaning that the speaker just randomly selects one of the messages without any optimality consideration. 
On the other hand, when $\lambda_\mathrm{S}\rightarrow \infty$, 
$\sigma(m \mid t)$ is non-zero only when $m$ yields the maximal utility, 
meaning that the speaker only selects the strictly optimal option(s).
This rule is widely used in the AI literature (e.g., \cite{Sutton1998RL}) and 
is also known as the \emph{exponential rule}.}.
The cost term is used to encode preference in
different utterances, be it about the utterances' lengths or syntactic
categories.

From (\ref{Bayesian_literal})-(\ref{Bayesian_sender_softmax}) we obtain the speaker's production rule\footnote{Note that we can further simplify (\ref{Bayesian_speaker}) as follows: $\exp(\lambda_\mathrm{S} \cdot (\log \mathcal{U}(t\mid \intp{m})-\mbox{Cost}(m)))
=\exp(\lambda_\mathrm{S} \log \mathcal{U}(t\mid \intp{m}) \cdot \exp(-\lambda_\mathrm{S}\cdot \mbox{Cost}(m))
=(\mathcal{U}(t\mid \intp{m}))^{\lambda_\mathrm{S}}\cdot \exp(-\lambda_\mathrm{S}\cdot \mbox{Cost}(m))$. Incidentally, if we do not take costs into account, i.e., let $\mbox{Cost}(m)\equiv 0$, then (\ref{Bayesian_speaker}) reduces to $\sigma(m \mid t) \propto (\mathcal{U}(t\mid \intp{m}))^{\lambda_\mathrm{S}}$, which is effectively a \emph{power rule} that soft-maximizes $\mathcal{U}(t\mid \intp{m})$. In this respect 
(\ref{Bayesian_speaker}) can be seen as an action-oriented speaker model (the exact definition of which will be given in the next section) equipped with 
a different kind of soft-max rule. However, we note that this is a mere technical coincidence, and as a first step, throughout this paper we only consider the exponential 
rule as the soft-max rule. \cite{Worthy2008ratio} provides a detailed discussion and empirical analyses on the differences between the exponential and power rules.}:

### 2014-03-14 03_alternatives.tex
minor corrections
several footnotes for reviewers' comments

The RSA model has a theoretical advantage over traditional formal
pragmatic theories in that it provides quantitative predictions, and
it has been shown to fit relevant empirical data well. However, if we
want to learn more about pragmatic reasoning about referential
expressions, it will be worthwhile to examine RSA carefully to pin
down its major components, to spell out its main design choices and
their underlying assumptions, and to test their contribution to the
predictive power of the model through statistical comparison with
natural alternatives. In this section, we will therefore compare the
RSA model to closely related game theoretic approaches, that likewise
assume that speakers and listeners form cascading beliefs about mutual behavior
and seek to optimize their behavior based on these beliefs
\cite{Rabin1990:Communication-b,Stalnaker:SayingMeaningCredibility,BenzvanRooijOptimalAssertions2007,Franke2011:Quantity-Implic,Jager2013:Rationalizable-}.\footnote{See
\cite{FrankeJager2013:Pragmatic-Back-} for overview and comparison
of game theoretic and Bayesian models.} Doing so
let's us describe a class of potential alternatives to RSA. We will
then compare the predictive power of these given relevant empirical
data in later sections.

@@

First of all, although it is prima facie reasonable to hypothesize, as
\cite{Frank} do, that the empirically measured perceptual salience
$\mathcal{S}(\cdot)$ is common knowledge between speaker and listener,
it does not actually affect RSA's production rule
(\ref{Bayesian_speaker}) at all. This makes it unclear in what sense
the speaker can be said to know the listener's perceptual salience,
since neither his beliefs nor actions depend on it. A natural variant,
which is used in some game theoretic models, would replace the literal
listener's uniform distribution $\mathcal{U}(t)$ in
(\ref{Bayesian_literal}) with the salience prior
$\mathcal{S}(t)$. This leads to the alternative production rule:
\begin{equation} \label{Bayesian_speaker_salience}
\sigma_\mathcal{S}(m \mid t) \propto \exp(\lambda_\mathrm{S} \cdot (\log \mathcal{S}(t\mid \intp{m})-\mbox{Cost}(m))) \enspace .
\end{equation}

Secondly, RSA measures the informativeness of an utterance $m$, which
is a crucial part of the communicative goal, in terms of how close the
induced belief of the literal listener is from the speaker's own
belief. This means that the RSA model sees the goal of communication
as conveying belief. While it is normally true that language does
convey the belief of the speaker, it is questionable at least in this
referential scenario whether letting the listener form a proper belief
is the ultimate goal of the communication. After all, if the speaker
wants to refer to something, it seems that in the end what matters is
whether the listener actually picks out the intended referent
successfully (e.g., when the speaker wants the listener to pass
something). This view is inherent in game theoretic approaches where
agents' beliefs are backed up by explicit formulations of their
utilities. We might call this latter view \emph{action-oriented}, in
contrast to the \emph{belief-oriented} view of communication which the
RSA model adopts, as it interprets the goal of communication as
invoking the intended action rather than forming an accurate
belief.\footnote{Note that the two views are clearly intimately
  related, despite our emphasis on their differences here. An accurate
  belief is usually helpful for choosing a desired action. In fact,
  forming a belief can itself be seen as an action that the listener
  performs, which might well be all that the speaker cares about in
  many everyday situations, e.g., casual talks or responses to
  inquiries of information. Also, the relevant actions might vary
  greatly across situations which makes belief formation a practical
  alternative to serve as the goal. Nevertheless, they are still
  essentially different views in spite of the fact that they often
  coincide in practice. The referential communication scenario we
  investigate in this paper reflects the distinction.} Thus, according
to this view, the informativeness of a message would be measured as
the probability of the listener choosing the intended
referent. Formally speaking, we have the action-oriented speaker model

@@

Hence we have four types of speaker models that differ in either the speaker's belief about the literal listener, or the speaker's goal of communication. We now introduce a uniform notation $\sigma_{xy},\ x\in\{a,b\},y\in\{\mathcal{U},\mathcal{S}\}$ for them:
\begin{equation} \label{Bayesian_speaker_uniform_action}
\sigma_{\mathrm{a}y}(m \mid t) \propto \exp(\lambda_\mathrm{S} \cdot ( y(t\mid m)-\mbox{Cost}(m))),
\end{equation}
\begin{equation} \label{Bayesian_speaker_uniform_belief}
\sigma_{\mathrm{b}y}(m \mid t) \propto \exp(\lambda_\mathrm{S} \cdot (\log y(t\mid m)-\mbox{Cost}(m))),
\end{equation}
where $\mathcal{U}$ is the uniform prior and $\mathcal{S}$ is the
salience prior. For example, in the original RSA model, the speaker
does not take listener's salience prior into account and he has a
belief-oriented goal of communication. Thus it will be denoted as
$\sigma_{\mathrm{b}\mathcal{U}}$.

>>>>

The RSA model has a theoretical advantage over traditional formal
pragmatic theories in that it provides quantitative predictions, and
it has been shown to fit relevant empirical data well. However, if we
want to learn more about pragmatic reasoning about referential
expressions, it will be worthwhile to examine RSA carefully to pin
down its major components, to spell out its main design choices and
their underlying assumptions, and to test their contribution to the
predictive power of the model through statistical comparison with
natural alternatives. In this section, we will therefore compare the
RSA model to closely related game theoretic approaches, that likewise
assume that speakers and listeners form cascading beliefs about mutual behavior
and seek to optimize their behavior based on these beliefs
\cite{Rabin1990:Communication-b,Stalnaker:SayingMeaningCredibility,BenzvanRooijOptimalAssertions2007,Franke2011:Quantity-Implic,Jager2013:Rationalizable-}.\footnote{See
\cite{FrankeJager2013:Pragmatic-Back-} for overview and comparison
of game theoretic and Bayesian models.} We will describe a class of potential 
alternatives to RSA and then compare the predictive power of these alternatives 
on relevant empirical data in later sections.

@@

First of all, although it is prima facie reasonable to hypothesize, as
\cite{Frank} do, that the empirically measured perceptual salience
$\mathcal{S}(\cdot)$ is common knowledge between speaker and listener,
it does not actually affect RSA's production rule
(\ref{Bayesian_speaker}) at all. This makes it unclear in what sense
the speaker can be said to know the listener's perceptual salience,
since neither his beliefs nor actions depend on it. A natural variant,
which is used in some game theoretic models, would be to replace the literal
listener's uniform distribution $\mathcal{U}(t)$ in
(\ref{Bayesian_literal}) with the salience prior
$\mathcal{S}(t)$. This leads to the alternative production rule:
\begin{equation} \label{Bayesian_speaker_salience}
\sigma_\mathcal{S}(m \mid t) \propto \exp(\lambda_\mathrm{S} \cdot (\log \mathcal{S}(t\mid \intp{m})-\mbox{Cost}(m))) \enspace .
\end{equation}

Secondly, RSA measures the informativeness of an utterance $m$, which
is a crucial part of the communicative goal, in terms of how close the
induced belief of the literal listener is from the speaker's own
belief. This means that the RSA model sees the goal of communication
as conveying belief. While it is normally true that language does
convey the belief of the speaker, it is questionable at least in this
referential scenario whether letting the listener form a proper belief
is the ultimate goal of the communication. After all, if the speaker
wants to refer to something, it seems that in the end what matters is
whether the listener actually picks out the intended referent
successfully (e.g., when the speaker wants the listener to pass
something). This view is inherent in game theoretic approaches where
agents' beliefs are backed up by explicit formulations of their
utilities. We might call this latter view \emph{action-oriented}, in
contrast to the \emph{belief-oriented} view of communication which the
RSA model adopts, as it interprets the goal of communication as
invoking the intended action rather than forming an accurate
belief.\footnote{Note that the two views are clearly intimately
  related, despite our emphasis on their differences here. An accurate
  belief is usually helpful for choosing a desired action. In fact,
  forming a belief can itself be seen as an action that the listener
  performs, which might well be all that the speaker cares about in
  many everyday situations, e.g., casual talks or responses to
  inquiries of information. Also, the relevant actions might vary
  greatly across situations which makes belief formation a practical
  alternative to serve as the goal. Nevertheless, they are still
  essentially different views in spite of the fact that they often
  coincide in practice. The referential communication scenario we
  investigate in this paper reflects the distinction.} Thus, according
to this view, the informativeness of a message would be measured as
the probability of the listener choosing the intended
referent. Formally speaking, we have the action-oriented speaker model\footnote{There might be concern about whether there should be another soft-max step for the literal listener's choice probability $\rho_0(t\mid m)$ in order for it to be action-oriented. 
This depends on the precise conceptual status of the literal listener in the model and to our knowledge there is no consensus yet. We tend to think of a literal listener as an automatic subconscious process that involves no optimality reasoning or decision-making. Thus we consider it appropriate to use $\rho_0(t\mid m)$ as the action of the literal listener. However, we do agree that in principle there can be a model whose literal listener does a soft-max over the salience prior, and such a model can be similarly included in the model comparison.}:

@@

Hence we have four types of speaker models that differ in either the speaker's belief about the literal listener, or the speaker's goal of communication. We now introduce a uniform notation $\sigma_{xy},\ x\in\{\mathrm{a},\mathrm{b}\},y\in\{\mathcal{U},\mathcal{S}\}$ for them:
\begin{equation} \label{Bayesian_speaker_uniform_action}
\sigma_{\mathrm{a}y}(m \mid t) \propto \exp(\lambda_\mathrm{S} \cdot ( y(t\mid m)-\mbox{Cost}(m))),
\end{equation}
\begin{equation} \label{Bayesian_speaker_uniform_belief}
\sigma_{\mathrm{b}y}(m \mid t) \propto \exp(\lambda_\mathrm{S} \cdot (\log y(t\mid m)-\mbox{Cost}(m))),
\end{equation}
where $\mathcal{U}$ is the uniform prior and $\mathcal{S}$ is the
salience prior. For example, in the original RSA model, the speaker
does not take listener's salience prior into account and he has a
belief-oriented goal of communication. Thus it will be denoted as
$\sigma_{\mathrm{b}\mathcal{U}}$.
